---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# inspercidados

<!-- badges: start -->
<!-- badges: end -->

**inspercidados** gives R users direct access to curated Brazilian urban
research datasets hosted on
[Insper's Dataverse](https://dataverse.datascience.insper.edu.br). The package
wraps [`dataverse`](https://github.com/IQSS/dataverse-client-r), so you can
find, download, and cite datasets with a few short commands.

## Installation

Install the development version from
[GitHub](https://github.com/portalcidados/inspercidados):

``` r
# install.packages("pak")
pak::pak("portalcidados/inspercidados")
```

## Core functions

| Function | Purpose |
|---|---|
| `list_datasets()` | List or search available datasets |
| `get_dataset()`   | Download a dataset into R |
| `cite_dataset()`  | Generate a citation for a dataset |
| `browse_project()` | Open the repository of the study behind a dataset |

## Browse available datasets

`list_datasets()` returns a tibble of every dataset in the registry. The
registry ships with the package, so no network call is needed.

```{r list}
library(inspercidados)

list_datasets()
```

Filter by alias, title, theme, region, or keywords:

```{r search}
list_datasets("Mobilidade")
```

## Download a dataset

Identify a dataset by its short alias, a bare DOI, or a full DOI URL:

```{r get, eval = FALSE}
# By alias
embarques <- get_dataset("embarques_mensais")

# By DOI
embarques <- get_dataset("10.60873/FK2/BPYHFB")

# Filter by year in multi-year datasets
pemob_2023 <- get_dataset("pemob_anual", year = 2023)

# Request a specific file or pattern
geo <- get_dataset("iptu_sp", filename = "iptu_2024.gpkg")
```

With `docs = TRUE`, the function returns the data together with its
documentation:

```{r docs, eval = FALSE}
result <- get_dataset("iptu_sp", docs = TRUE)
result$data
result$docs
```

## Cite a dataset

`cite_dataset()` fetches metadata from Dataverse and returns a citation as
plain text, BibTeX, or RIS:

```{r cite, eval = FALSE}
cite_dataset("embarques_mensais")
cite_dataset("embarques_mensais", format = "bibtex")
cite_dataset("embarques_mensais", format = "ris")
```

## Explore the study behind a dataset

Most datasets come from a research study whose pipeline lives in its own
repository. `browse_project()` prints the study, lists its datasets, and opens
the repository:

```{r project, eval = FALSE}
browse_project("embarques_mensais")
```

## Learn more

- Full reference: <https://portalcidados.github.io/inspercidados>
- Data source: <https://dataverse.datascience.insper.edu.br>
