---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Curated datasets from Insper Cidades

<!-- badges: start -->
```{r, echo = FALSE, results = 'asis', eval = requireNamespace("badger", quietly = TRUE)}
cat(badger::badge_runiverse())
```
[![R-CMD-check](https://github.com/inspercidades/inspercidados/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/inspercidades/inspercidados/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**inspercidados** gives R users direct access to datasets from Insper Cidades,
hosted on [Insper Dataverse](https://dataverse.datascience.insper.edu.br).
Insper Cidades is a multidisciplinary research center at Insper. The package
wraps [`dataverse`](https://github.com/IQSS/dataverse-client-r) so you can
find, download, and cite datasets in a few short commands.

## Installation

The stable version of the package is available on CRAN.

``` r
install.packages("inspercidados")
```

The development version of the package is available on
[R-universe](https://inspercidades.r-universe.dev/inspercidados).

``` r
install.packages(
  "inspercidados",
  repos = c(
    "https://inspercidades.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

## Core functions

| Function | Purpose |
|---|---|
| `list_datasets()` | List or search available datasets |
| `list_resources()` | List logical resources within a dataset |
| `get_dataset()`   | Download a dataset into R |
| `get_dataverse()` | Inspect or download any Insper Dataverse deposit |
| `cite_dataset()`  | Generate a citation for a dataset |
| `open_project()` | Open the study repository behind a dataset |

## Browse available datasets

`list_datasets()` returns a tibble of every dataset in the registry. The
registry ships with the package, so no network call is needed.

```{r list}
library(inspercidados)

list_datasets()
```

Search by alias, title, description, theme, region, or keywords.

```{r search}
list_datasets("Mobilidade")
```

## Download a dataset

Pass a short alias to download a registered dataset.

```{r get, eval = FALSE}
# By alias
embarques <- get_dataset("embarques_mensais")

# Filter by year in multi-year datasets
pemob_2023 <- get_dataset("pemob_anual", year = 2023)

# Select a logical resource and format
pontos <- get_dataset(
  "qualidade_ar_mare",
  resource = "pontos",
  format = "gpkg"
)
```

`list_resources()` shows the logical datasets and formats inside a
registered deposit.

```{r resources}
list_resources("qualidade_ar_mare")
```

Set `docs = TRUE` to return the data alongside its documentation.

```{r docs, eval = FALSE}
result <- get_dataset("iptu_sp", docs = TRUE)
result$data
result$docs
```

## Cite a dataset

`cite_dataset()` fetches metadata from Dataverse and returns a citation in
plain text, BibTeX, or RIS.

```{r cite, eval = FALSE}
cite_dataset("embarques_mensais")
cite_dataset("embarques_mensais", format = "bibtex")
cite_dataset("embarques_mensais", format = "ris")
```

## Explore the study behind a dataset

Most datasets come from a research study whose pipeline lives in its own
repository. `open_project()` prints the study, lists its datasets, and opens
the repository.

```{r project, eval = FALSE}
open_project("faixa_azul")
```

## Learn more

To learn more about the package, check <https://inspercidades.github.io/inspercidados/>
