---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# inspercidados

<!-- badges: start -->
[![R-CMD-check](https://github.com/inspercidades/inspercidados/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/inspercidades/inspercidados/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/inspercidades/inspercidados/graph/badge.svg)](https://app.codecov.io/gh/inspercidades/inspercidados)
<!-- badges: end -->

**inspercidados** provides simple, reproducible access to curated Brazilian
urban research datasets hosted on
[Insper's Dataverse](https://dataverse.datascience.insper.edu.br). It is a
lightweight wrapper around the
[`dataverse`](https://github.com/IQSS/dataverse-client-r) package that lets
researchers discover, download, and cite datasets with a few short commands.

## Installation

You can install the development version of inspercidados from
[GitHub](https://github.com/portalcidados/inspercidados) with:

``` r
# install.packages("pak")
pak::pak("portalcidados/inspercidados")
```

## Core functions

| Function | Purpose |
|---|---|
| `list_datasets()` | List or search available datasets |
| `get_dataset()`   | Download a dataset into R |
| `cite_dataset()`  | Generate a citation for a dataset |
| `get_script()`    | Open a companion analysis or replication script |

## Browse available datasets

`list_datasets()` returns a tibble of all datasets in the package registry.
No network call is made.

```{r list}
library(inspercidados)

list_datasets()
```

You can filter by alias, title, theme, region, or keywords:

```{r search}
list_datasets("Mobilidade")
```

## Download a dataset

Datasets can be identified by their short alias, a bare DOI, or a full DOI URL:

```{r get, eval = FALSE}
# By alias
embarques <- get_dataset("embarques_mensais")

# By DOI
embarques <- get_dataset("10.60873/FK2/BPYHFB")

# Filter by year for multi-year datasets
pemob_2023 <- get_dataset("pemob_anual", year = 2023)

# Request a specific file or pattern
geo <- get_dataset("iptu_sp", filename = "iptu_2024.gpkg")
```

Pass `docs = TRUE` to return the dataset together with its documentation:

```{r docs, eval = FALSE}
result <- get_dataset("iptu_sp", docs = TRUE)
result$data
result$docs
```

## Cite a dataset

`cite_dataset()` fetches metadata from Dataverse and returns a citation in
plain text (default), BibTeX, or RIS:

```{r cite, eval = FALSE}
cite_dataset("embarques_mensais")
cite_dataset("embarques_mensais", format = "bibtex")
cite_dataset("embarques_mensais", format = "ris")
```

## Companion R scripts

Some datasets ship with a companion script that demonstrates how to load and
explore the data, or documents the production pipeline that generated it:

```{r script, eval = FALSE}
# Open the analysis script in your editor
get_script("embarques_mensais")

# Open the replication pipeline (typically not runnable by external users)
get_script("embarques_mensais", type = "replication")
```

## Learn more

- `vignette("getting-started", package = "inspercidados")`
- Full reference: <https://portalcidados.github.io/inspercidados>
- Data source: <https://dataverse.datascience.insper.edu.br>
