---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,
  purl     = FALSE
)
```

`inspercidados` gives you the curated Brazilian urban datasets published by
[Insper Cidades](https://www.insper.edu.br) on Insper's Dataverse. The package
wraps the [`dataverse`](https://CRAN.R-project.org/package=dataverse) package
and adds a registry that maps short names to deposits.

## Installation

```{r install}
remotes::install_github("inspercidades/inspercidados")
```

## Finding a dataset

`list_datasets()` reads the local registry, so it answers instantly and works
offline.

```{r list}
library(inspercidados)

# Everything you can download
list_datasets()

# Search titles, descriptions, keywords, themes and regions
list_datasets("Mobilidade")

# Narrow by column
list_datasets(theme = "Habitação")
list_datasets(spatial = TRUE)
list_datasets(project = "faixa_azul")
```

Each row carries an `alias`, which every other function accepts. The `formats`
column reports the file types deposited, and `is_spatial` marks the datasets
that return an `sf` object.

Some datasets appear in the catalog but are not published for download, because
they are held in Insper's secure data room. `include = "catalogued"` shows them
with `access = "secure_room"`.

```{r list-secure}
list_datasets(include = "catalogued", search = "iFood")
```

## Downloading a dataset

`get_dataset()` accepts an alias, a bare DOI, or a DOI URL.

```{r download}
embarques <- get_dataset("embarques_mensais")

# The same deposit, by DOI
embarques <- get_dataset("10.60873/FK2/BPYHFB")

# Or by URL
embarques <- get_dataset("https://doi.org/10.60873/FK2/BPYHFB")
```

Deposits usually hold the same data in several formats. `get_dataset()` picks
one for you, favouring GeoPackage for spatial datasets and otherwise RDS,
Parquet, delimited text, and Excel in that order. Formats that need a suggested
package are skipped when it is not installed.

Three arguments override the choice.

```{r download-filters}
# Pick by year, matched against the filename
pemob_2023 <- get_dataset("pemob_anual", year = 2023)

# Pick an exact file
iptu_geo <- get_dataset("iptu_sp", filename = "iptu_residencial.gpkg")

# Pick by pattern
iptu_geo <- get_dataset("iptu_sp", file_pattern = "\\.gpkg$")
```

### Aliases that share a deposit

One Dataverse deposit can hold several datasets. The PEMOB deposit carries both
the annual series and the harmonised panel, so `"pemob_anual"` and
`"pemob_harmonizada"` resolve to the same DOI and differ by a file pattern the
registry stores for you.

```{r shared}
anual <- get_dataset("pemob_anual", year = 2024)
painel <- get_dataset("pemob_harmonizada")
```

## Documentation alongside the data

`docs = TRUE` returns a list with the data and its documentation. A
documentation workbook deposited with the data takes precedence; otherwise the
function falls back to the Dataverse metadata record.

```{r docs}
result <- get_dataset("embarques_mensais", docs = TRUE)
result$data
result$docs
```

## Spatial datasets

Datasets flagged `is_spatial` resolve to a GeoPackage and come back as `sf`
objects. Without `sf` installed, `get_dataset()` warns and falls back to a
tabular format.

```{r spatial}
library(sf)

faixa <- get_dataset("faixa_azul_sp")
plot(st_geometry(faixa))
```

## Finding the code behind a dataset

Each dataset comes from a research study, and a study normally produces several
datasets from one pipeline. Pipelines are therefore published as repositories
rather than as a script per dataset.

```{r projects}
# Every study, its repository, and the datasets it produced
list_projects()

# The study behind one dataset
browse_project("sinistros_sp")
```

`browse_project()` prints the study, its sibling datasets, and its repository,
then opens the repository in your browser. Repositories marked `private` in
`list_projects()` need an Insper Cidades GitHub account, and some studies have
not published their code yet.

## Datasets outside the registry

`get_dataverse()` downloads any Insper Dataverse deposit, including those with
no alias. Paste a DOI or a landing-page URL straight from the browser.

```{r dataverse}
# Inspect a deposit before downloading
get_dataverse("10.60873/FK2/AOLEOI", files = TRUE)

# Then download it
d <- get_dataverse("10.60873/FK2/AOLEOI")
```

Without registry knowledge, this function guesses which file you want from the
format preference alone. Deposits with an unusual layout may need `filename` or
`file_pattern`, and it serves only DOIs under the `10.60873` prefix.

## Citing a dataset

```{r cite}
cite_dataset("embarques_mensais")
cite_dataset("embarques_mensais", format = "bibtex")
cite_dataset("embarques_mensais", format = "ris")
```
