---
title: "Getting started with inspercidados"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with inspercidados}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(inspercidados)
```

inspercidados gives R users direct access to curated Brazilian urban research
datasets hosted on [Insper's Dataverse](https://dataverse.datascience.insper.edu.br).
It is a thin wrapper around the [`dataverse`](https://github.com/IQSS/dataverse-client-r)
package.

## Open science

The package supports open and reproducible research in three ways:

- **Findable and accessible data.** Urban research data for Brazil is scattered
  across portals and formats. The package registers each dataset with a stable
  alias and a DOI, so anyone can find and download the same data with one call.
- **Proper attribution.** Published research often overlooks the datasets
  behind its results. `cite_dataset()` makes citing the data as easy as citing
  a paper.
- **Transparency of methods.** Datasets come from research projects whose
  pipelines are published as repositories. `browse_project()` takes you from a
  dataset to the code that produced it.

## Installation

``` r
# install.packages("pak")
pak::pak("inspercidades/inspercidados")
```

## List datasets

`list_datasets()` reads the registry that ships with the package; it makes no
network calls:

```{r list}
list_datasets()
```

Search by title, theme, region, or keywords:

```{r search}
list_datasets("Mobilidade")
```

## Download a dataset

Identify a dataset by its short alias, a bare DOI, or a full DOI URL:

```{r download, eval = FALSE}
# By alias
embarques <- get_dataset("embarques_mensais")

# Filter by year in multi-year datasets
pemob_2023 <- get_dataset("pemob_anual", year = 2023)
```

## Cite a dataset

```{r cite, eval = FALSE}
cite_dataset("embarques_mensais")
```

## Next steps

The full reference is at <https://inspercidades.github.io/inspercidados>.
