---
title: "Getting started with inspercidados"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with inspercidados}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(inspercidados)
```

`inspercidados` gives users direct access to curated datasets hosted on [Insper's Dataverse](https://dataverse.datascience.insper.edu.br). These datasets come from research publications made by authors related to [Insper Cidades](https://www.insper.edu.br/pt/pesquisa/centro-de-estudos-das-cidades), and that are featured as data visualization narratives on the [Cidados website](https://cidados.insper.edu.br).

## Open science

This package is a core element in a suite of open science initiatives from Insper. It supports open and reproducible research in two ways:

- **Findable and accessible open data.** The package registers each dataset with a stable
  alias and a DOI, so anyone can find and download the same data with one call.
- **Transparency and replication.** Datasets come from research projects whose
  pipelines are published as repositories. `open_project()` takes you from a
  dataset to the code that produced it.

## Installation

The package is available on R-universe:

```r
install.packages(
  "inspercidados",
  repos = c(
    "https://inspercidades.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

The development version is available on GitHub.

``` r
# install.packages("pak")
pak::pak("inspercidades/inspercidados")
```

## Using the package

`list_datasets()` reads the registry that ships with the package; it makes no
network calls.

```{r list}
list_datasets()
```

To search by title, theme, region, or keywords use the `search` argument in `list_dataset()`

```{r search}
list_datasets(search = "Mobilidade")
```

Each registered dataset has a short alias. Use `get_dataset()` to download it.

```{r download, eval = FALSE}
# By alias
embarques <- get_dataset("embarques_mensais")

# Filter by year in multi-year datasets
pemob_2023 <- get_dataset("pemob_anual", year = 2023)

# Select one logical resource and file format
pontos <- get_dataset(
  "qualidade_ar_mare",
  resource = "pontos",
  format = "gpkg"
)
```

Some Dataverse deposits contain several logical datasets. Use
`list_resources()` to inspect them without making a network request.

```{r resources}
list_resources("qualidade_ar_mare")
```

## Cite a dataset

```{r cite, eval = FALSE}
cite_dataset("embarques_mensais")
```

## Next steps

The full reference is at <https://inspercidades.github.io/inspercidados>.
