---
title: "Mapping maternal mortality in São Paulo"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
```

```{r setup}
library(inspercidados)
library(dplyr)
library(sf)
library(ggplot2)
```

The `mortalidade_sp` deposit pairs São Paulo's district boundaries with health
indicators for the city's 96 districts. This article downloads the boundaries
and the maternal-mortality table, joins them by district code, and maps the
maternal mortality ratio. The full narrative is on the
[Cidados website](https://cidados.insper.edu.br/historias/desigualdades-em-saude-sp).

## Download the layers

`get_dataset()` returns an `sf` object for spatial resources, so the boundary
layer arrives ready to map.

```{r boundaries}
distritos <- get_dataset("mortalidade_sp", format = "gpkg")
distritos
```

The maternal-mortality annex holds one row per district, year, and indicator.

```{r maternal}
materna <- get_dataset(
  "mortalidade_sp",
  resource = "mortalidade_materna"
)
glimpse(materna)
```

## Build the ratio

Pool the decade to steady the ratio, then express it as deaths per 100,000 live
births.

```{r ratio}
razao <- materna |>
  filter(indicador %in% c("obitos_maternos", "nascidos_vivos")) |>
  summarise(
    obitos = sum(valor[indicador == "obitos_maternos"], na.rm = TRUE),
    nascidos = sum(valor[indicador == "nascidos_vivos"], na.rm = TRUE),
    .by = distrito
  ) |>
  mutate(razao = obitos / nascidos * 1e5)
```

Join the ratio onto the boundaries by `distrito`.

```{r join}
mapa <- distritos |>
  left_join(razao, by = "distrito")
```

## Map the districts

```{r map, fig.alt = "Choropleth map of the maternal mortality ratio by district in São Paulo, ranging from zero to about 100 deaths per 100,000 live births."}
ggplot(mapa) +
  geom_sf(aes(fill = razao), colour = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colours = RColorBrewer::brewer.pal(9, "YlOrRd"),
    na.value = "grey90",
    name = "Deaths per\n100,000 live births"
  ) +
  labs(
    title = "Maternal mortality by district",
    subtitle = "São Paulo, 2010–2019"
  ) +
  theme_void()
```

## Learn more

- [The story behind the data](https://cidados.insper.edu.br/historias/desigualdades-em-saude-sp)
- [Full package reference](https://inspercidades.github.io/inspercidados)
